Add Data and Data Dictionary

### 2025-07-15 data.qmd
Shorten the data chapter

---
title: "Add Data"
engine: knitr
---

@@

invisible(loadNamespace("palmerpenguins")) # Tell renv that we need this package
stopifnot(packageDescription("palmerpenguins", fields = "License") == "CC0")
write.csv(palmerpenguins::penguins, file = "data.csv", row.names = FALSE)
```

@@

[data.csv](data.csv){.btn .btn-lg .btn-info download=""}

@@

## Add Data Dictionary

Whether or not distributing the data set,
it is important to document the meaning (e.g., units)
and values of its variables.
This is typically done with a _data dictionary_ (also called a _codebook_).
In the following, we will demonstrate how to create a simple data dictionary
using the R package [`datawizard`][datawizard].
You can install them now using:

[datawizard]: https://easystats.github.io/datawizard/

```{.r filename="Console"}
renv::install("datawizard")
```

You can put the code that follows in a separate document.
Create it by clicking on _File_ > _New File_  > _Quarto Document..._.
Choose a title such as `Data Dictionary`,
select _HTML_ as format,
uncheck the use of the visual markdown editor, and click on _Create_.
Remove everything except the YAML header (between the `---`).
To make the HTML file self-contained,
also set `embed-resources: true` such that the YAML header looks as follows:

```{.yml filename="data_dictionary.qmd"}
---
title: "Data Dictionary"
format:
  html:
    embed-resources: true
---
```

Then, save it as `data_dictionary.qmd` by clicking on _File_ > _Save_.

To create the actual data dictionary, first write a description for all columns
so others can understand what the variable names mean.
Where necessary, also document their value
-- this is especially important if their meaning is non-obvious.
In the following, we demonstrate this by storing the penguins' binomial name
along with the English name.

``````{cat}
#| engine.opts: { file: "_data_dictionary.qmd" }
#| class.source: "md"
#| filename: "data_dictionary.qmd"

```{r}
#| echo: false

# Store the description of variables
vars <- c(
  species = "a character string denoting penguin species",
  island = "a character string denoting island in Palmer Archipelago, Antarctica",
  bill_length_mm = "a number denoting bill length (millimeters)",
  bill_depth_mm = "a number denoting bill depth (millimeters)",
  flipper_length_mm = "an integer denoting flipper length (millimeters)",
  body_mass_g = "an integer denoting body mass (grams)",
  sex = "a character string denoting penguin sex",
  year = "an integer denoting the study year"
)

# Store the description of variable values
vals <- list(
  species = c(
    Adelie = "Pygoscelis adeliae",
    Gentoo = "Pygoscelis papua",
    Chinstrap = "Pygoscelis antarcticus"
  )
)
```
``````

Then, load the data and use `datawizard`
to add the descriptions to the `data.frame`:[^not-permanent]

[^not-permanent]: Note that the code provided does not alter the data file
-- no description will be added to `data.csv`.
The descriptions are only added to a (temporary) copy of the data set within R
to create the data dictionary.

::: {.column-margin}
![datawizard: Easy Data Wrangling and Statistical Transformations](images/datawizard.png){width=250px}
:::

``````{cat}
#| engine.opts: { file: "_data_dictionary.qmd", append: TRUE }
#| class.source: "md"
#| filename: "data_dictionary.qmd"

```{r}
#| echo: false

dat <- read.csv("data.csv")

for (x in names(vars)) {
  if (x %in% names(vals)) {
    dat <- datawizard::assign_labels(
      dat,
      select = I(x),
      variable = vars[[x]],
      values = vals[[x]]
    )
  } else {
    dat <- datawizard::assign_labels(
      dat,
      select = I(x),
      variable = vars[[x]]
    )
  }
}
```

``````

Then, you can create the data dictionary containing the descriptions,
but also some other information about each variable
(e.g., the number of missing values) and print it.

``````{cat}
#| engine.opts: { file: "_data_dictionary.qmd", append: TRUE }
#| class.source: "md"
#| filename: "data_dictionary.qmd"

```{r}
#| echo: false
#| column: "body-outset"
#| classes: plain

datawizard::data_codebook(dat) |>
  datawizard::data_select(exclude = ID) |>
  datawizard::data_filter(N != "") |>
  datawizard::print_md()
```

``````

```{r}
#| child: "_data_dictionary.qmd"

```

Depending on the type of data, it may also be necessary
to describe sampling procedures (e.g., selection criteria),
measurement instruments (e.g., questionnaires),
appropriate weighting,
already applied preprocessing steps, or contact information.
In our case, as the data has already been published,
we only store a reference to its source.

The data set is from the R package `palmerpenguins`.
If you had it installed
you could use the function `citation()` to create such a reference:

```{r}
#| label: "data-citation"
#| eval: false

citation("palmerpenguins", auto = TRUE) |>
  format(bibtex = FALSE, style = "text")
```

Without the package `palmerpenguins` installed,
you can find a [suggested citation on its website][palmerpenguins-citation]
and add that to your data dictionary:

[palmerpenguins-citation]: https://allisonhorst.github.io/palmerpenguins/#citation

```{r}
#| ref.label = "data-citation",
#| render = function(x, options) gsub("\\n", " ", x = x),
#| echo = FALSE,
#| class.output = "md code-overflow-wrap",
#| attr.output = 'filename="data_dictionary.qmd"'

# This chunk takes the output from the chunk "data-citation"
# and renders it with all newlines replaced by whitespaces.
```

Finally, you can render the data dictionary by running the following:

```{.bash filename="Terminal"}
quarto render data_dictionary.qmd
```

This should create the file `data_dictionary.html`
which you open and view in your web browser.

One could go even further by making the information machine-readable in a standardized way.
We provide an optional example of that in @nte-frictionless.
If you want to learn more about the sharing of research data,
have a look at the tutorial "[FAIR research data management][fair-tutorial]".

[fair-tutorial]: https://lmu-osc.github.io/FAIR-Data-Management/

::: {#nte-frictionless .callout-note collapse="true"}
### Create Machine-Readable Variable Documentation

This example demonstrates how the title and description of the data set,
the description of the variables and their valid values are stored in a machine-readable way.
We'll reuse the descriptions we already created[^value-labels] and add a few others.

[^value-labels]: Unfortunately, the descriptions of values are not reused in this example,
as they are [not supported][enum-labels] by the specification we are using.

[enum-labels]: https://specs.frictionlessdata.io/patterns/#table-schema-enum-labels-and-ordering

First, store the title and description of the data set as a whole:

```{.r filename="Console"}
table_info <- c(
  title = "penguins data set",
  description = "Size measurements for adult foraging penguins near Palmer Station, Antarctica"
)
```

As before, also provide a reference to the source.

```{r}
#| echo: false
#| class-output: "r code-overflow-wrap"
#| attr-output: 'filename="Console"'

# We have provided the data set as CSV file to the readers.
# Therefore, we cannot assume or require that readers have
# the R package palmerpenguins installed. Instead, we create
# the citation on our end and hide how we obtained it.

citation("palmerpenguins", auto = TRUE)$url |>
  paste0("dat_source <- \"", ... = _, "\"") |>
  cat()
```

Next, create a list of the categorical variables' valid values:

```{.r filename="Console"}
valid_vals <- list(
  species = c("Adelie", "Gentoo", "Chinstrap"),
  island = c("Torgersen", "Biscoe", "Dream"),
  sex = c("male", "female"),
  year = c(2007, 2008, 2009)
)
```

Finally, store the descriptions of the variables we already created earlier:

```{.r filename="Console"}
# Store the description of variables
vars <- c(
  species = "a character string denoting penguin species",
  island = "a character string denoting island in Palmer Archipelago, Antarctica",
  bill_length_mm = "a number denoting bill length (millimeters)",
  bill_depth_mm = "a number denoting bill depth (millimeters)",
  flipper_length_mm = "an integer denoting flipper length (millimeters)",
  body_mass_g = "an integer denoting body mass (grams)",
  sex = "a character string denoting penguin sex",
  year = "an integer denoting the study year"
)
```

Generally, metadata are either stored embedded into the data or externally,
for example, in a separate file.
We will use the "[frictionless data](https://frictionlessdata.io/)" standard,
where metadata are stored separately.
Another alternative would be [RO-Crate](https://www.researchobject.org/ro-crate/).

Specifically, one can use the R package [`frictionless`][frictionless]
to create a _schema_ which describes the structure of the data.[^frictionless-note]
For the purpose of the following code,
it is just a nested list that we edit to include our own information.
We also explicitly record in the schema
that missing values are stored in the data file as `NA`
and that the data are licensed under CC0\ 1.0.
Finally, the package is used to create a metadata file that contains the schema.

[frictionless]: https://docs.ropensci.org/frictionless/

[^frictionless-note]: In June 2024, [version 2](https://datapackage.org/)
of the frictionless data standard has been released.
As of November 2024, the R package `frictionless` only supports the first version,
though support for v2 is [planned](https://github.com/frictionlessdata/frictionless-r/labels/datapackage%3Av2).

```{.r filename="Console"}
# Install {frictionless} and the required dependency {stringi}
renv::install(c(
  "frictionless",
  "stringi"
))

# Read data and create schema
dat_filename <- "data.csv"
dat <- read.csv(dat_filename)
dat_schema <- frictionless::create_schema(dat)

# Add descriptions to the fields
dat_schema$fields <- lapply(dat_schema$fields, \(x) {
  c(x, description = vars[[x$name]])
})

# Record valid values
dat_schema$fields <- lapply(dat_schema$fields, \(x) {
  if (x[["name"]] %in% names(valid_vals)) {
    modifyList(x, list(constraints = list(enum = valid_vals[[x$name]])))
  } else {
    x
  }
})

# Define missing values
dat_schema$missingValues <- c("", "NA")

# Create package with license info and write it
dat_package <- frictionless::create_package() |>
  frictionless::add_resource(
    resource_name = "penguins",
    data = dat_filename,
    schema = dat_schema,
    title = table_info[["title"]],
    description = table_info[["description"]],
    licenses = list(list(
      name = "CC0-1.0",
      path = "https://creativecommons.org/publicdomain/zero/1.0/",
      title = "CC0 1.0 Universal"
    )),
    sources = list(list(
      title = "CRAN",
      path = dat_source
    ))
  )
frictionless::write_package(dat_package, directory = ".")
```

This creates the metadata file `datapackage.json` in the current directory.
Make sure it is located in the same folder as `data.csv`,
as together they comprise a [data package](https://specs.frictionlessdata.io/data-package/).
:::

Having added the data and its documentation,
one can view and record the utilized packages with `renv`,
thus bringing the project into a consistent state:

```{.r filename="Console"}
renv::status()
renv::snapshot()
```

## Add Data Citation and Attribution

All data relied upon should be cited in the manuscript
to allow for precise identification and access.
From the "eight core principles of data citation" by @Starr2015,
licensed under [CC0\ 1.0][cc0]:

> **Principle 1 – Importance**: "Data should be considered legitimate,
> citable products of research. Data citations should be accorded the same
> importance in the scholarly record as citations of other research objects,
> such as publications."
> 
> **Principle 3 – Evidence**: "In scholarly literature, whenever and wherever a
> claim relies upon data, the corresponding data should be cited."
> 
> **Principle 5 – Access**: "Data citations should facilitate access to the data
> themselves and to such associated metadata, documentation, code, and other
> materials, as are necessary for both humans and machines to make informed use
> of the referenced data."
> 
> **Principle 7 – Specificity and Verifiability**: "Data citations should
> facilitate identification of, access to, and verification of the specific data
> that support a claim. Citations or citation metadata should include
> information about provenance and fixity sufficient to facilitate verifying
> that the specific time slice, version and/or granular portion of data
> retrieved subsequently is the same as was originally cited."

Now, it's your turn to add an appropriate citation for the data set to the manuscript.
Does your citation adhere to the principles above?

@@

You can find an appropriate BibTeX entry on [the package website][palmerpenguins-citation]
or with the function `citation()`:^[Note that this function requires
to have the respective package installed.]

@@

While citation happens in the manuscript for reasons of academic integrity and reproducibility,
to comply with any licenses you also may need to provide attribution within your project folder.
Even though the data file we use here does not require attribution,
we recommend adding a short paragraph to `LICENSE.txt`:

```{r}
#| echo: false
#| class-output: "txt code-overflow-wrap"
#| attr-output: 'filename="LICENSE.txt"'

cat(paste0(
  "The penguins data stored in \"data.csv\" by Allison Horst, Alison Hill, and Kristen Gorman available from <",
  citation("palmerpenguins", auto = TRUE)$url,
  "> are licensed under CC0 1.0: <https://creativecommons.org/publicdomain/zero/1.0/>"
))
```

As before, if the license required adding the full license text,
you would also need to copy it to the project folder (if not already in there).

Finally, you can go through the commit routine:

>>>>

---
title: "Add Data and Data Dictionary"
engine: knitr
---

@@

invisible(loadNamespace("palmerpenguins")) # Tell renv that we need this package
stopifnot(packageDescription("palmerpenguins", fields = "License") == "CC0")

# write both in the top folder and in the in-depth folder
# (so that both scripts can reference the data without relative paths)
write.csv(palmerpenguins::penguins, file = "data.csv", row.names = FALSE)
write.csv(palmerpenguins::penguins, file = "in_depth_material/data.csv", row.names = FALSE)
```

@@

[data.csv](data.csv){.btn .btn-lg .btn-info download="data.csv"}

@@

## Add a Data Dictionary

Whether or not distributing the data set,
it is important to document the meaning (e.g., units)
and values of its variables.
This is typically done with a _data dictionary_ (also called a _codebook_).

The recommendation for data dictionaries vary between fields - 
both in terms of the recommended content (i.e., what exactly should be documented) 
and the technical implementation (i.e., which file formats should be used).

*For the purpose of this exercise, we keep it easy and propose that you 
manually create a file with the data dictionary
(e.g., as a table in .xlsx, .docx, .ods, or as a Markdown table), 
documenting only the bare minimum.*

<!-- TODO: Add a link to the forthcoming data dictionary OSC tutorial -->

::: {#tip-in-depth .callout-tip collapse="false"}
### Manual is too much work? Automatic generation of (machine-readable) data dictionaries
Our in-depth supplementary material "[Automatic Generation of Data Dictionaries](in_depth_material/data_dic_generation.qmd)" explains how you can automatically create
a data dictionary with an R package. The package reads the data set and extracts relevant information from it. 
This approach is in particular useful if you need to document data sets with many variables.

This advanced chapter also contains a section on how to create machine-readable data dictionaries.
:::

### A bare minimum data dictionary

Most standards for data dictionaries require at least 
this information for each variable in your data set:

- `name`: The (machine‑readable) name of the variable  
- `label`: A short, human‑readable title or label for the variable  
- `type`: The data type of the variable (e.g., `integer`, `float`, `string`, `date`)  
- `description`: A brief description of what the variable measures or represents  
- `values` (for categorical variables): A mapping of codes to their meanings, for example:  1 = Male, 2 = Female, 9 = Missing)
- `units` (for numeric measures): The units of measurement (e.g., `kg`, `USD`, `years`, or the scale of a survey response), 
- `missing_codes`: Any special codes used to denote missing or non‑applicable values (e.g., `-99 = Not answered`)

Here's an example from a different data set, with variables in rows, and the dictionary in columns:

| name              | label                     | type    | description                             | values                                                                                  | units       | missing_codes      |
| ----------------- | ------------------------- | ------- | --------------------------------------- | --------------------------------------------------------------------------------------- | ----------- | ------------------ |
| gender            | Gender                    | integer | self‑identified gender                  | 1 = Male; 2 = Female; 3 = Other; 9 = Missing                                            |             | 9 = Missing        |
| age               | Age                       | integer | age in years                            |                                                                                         | years       | -99 = Not answered |
| blood_pressure    | Blood Pressure (systolic) | integer | systolic blood pressure                 |                                                                                         | mmHg        | -99 = Not measured |
| life_satisfaction | Life Satisfaction         | integer | "How satisfied are you with your life?" | 1 = Very dissatisfied; 2 = Dissatisfied; 3 = Neutral; 4 = Satisfied; 5 = Very satisfied | scale (1–5) | -99 = Not answered |

### ✍️ Practical Exercise: Add your own data dictionary

Now go ahead and create a data dictionary for the penguins data set, 
in a software (text or spreadsheet) of your choice.

Save the data dictionary in the same folder as the actual data set file.

::: {#tip-data_dic .callout-tip collapse="true"}
#### Data dictionary for penguins data set (Solution)

| name              | label          | type    | description                          | values                    | units | missing_codes |
| ----------------- | -------------- | ------- | ------------------------------------ | ------------------------- | ----- | ------------- |
| species           | Species        | string  | Penguin species                      | Adelie; Gentoo; Chinstrap |       | NA = Missing  |
| island            | Island         | string  | Island where individual was observed | Torgersen; Biscoe; Dream  |       | NA = Missing  |
| bill_length_mm    | Bill length    | float   | Length of the bill (beak)            |                           | mm    | NA = Missing  |
| bill_depth_mm     | Bill depth     | float   | Depth of the bill (beak)             |                           | mm    | NA = Missing  |
| flipper_length_mm | Flipper length | integer | Length of the flipper                |                           | mm    | NA = Missing  |
| body_mass_g       | Body mass      | integer | Body mass                            |                           | g     | NA = Missing  |
| sex               | Sex            | string  | Sex of the penguin                   | male; female              |       | NA = Missing  |
| year              | Year           | integer | Year of data collection              | 2007; 2008; 2009          |       |               |

:::

## ✍️ Practical Exercise: Add Data Citation and Attribution

All data relied upon should be cited in the manuscript
to allow for precise identification and access.
Now, it's your turn to add an appropriate citation for the data set to the manuscript.

Hints:

- You can find an appropriate BibTeX entry on [the package website](https://allisonhorst.github.io/palmerpenguins/#citation)
  or with the function `citation()`^[Note that this function requires to have the respective package installed.].
- Add the citation in the manuscript where it says "cite data here".

@@

Show the correct reference of the data set:

@@

## Wrap up

Congrats! You documented your data set and cited it correctly.

To finalize this step, you can go through the commit routine:
